---
title: "Biomart"
author: "Marc Galland"
date: "21 septembre 2016"
output: html_document
---
BiomaRt is a R package to retrieve data from BioMart databases.
From the documentation: 
> It enables to retrieve large amounts of data in a uniform way without the need to know the underlying database schemas or write complex SQL queries. Some BioMart databases are Ensembl, Uniprot and HapMap. 

```{r}
library(biomaRt)
```

# Biomart

## connect to a BioMart database and list available datasets
It first extract a database ("Mart") for *homo sapiens*
```{r}
# get human Mart 
humanMart = useMart("ensembl",dataset = "hsapiens_gene_ensembl")
```

## Query selected biomart database
__Filters__: restriction on the datasets. Filters (one or more) that should be used in the query. To see all possible filters, type `View(filters)`
__Attributes__: values we want to retrieve. Attributes you want to retrieve. To see all possible attributes, type `View(attributes)`
```{r BiomaRt query}
# All possible filters
filters = humanMart@filters

# All possible atttributes
attributes = humanMart@attributes

# # build the query for a specific gene
# humanTrans = getBM(
#   # values we want to retrieve
#   attributes = c('entrezgene','hgnc_symbol','ensembl_gene_id','affy_hugene_1_0_st_v1'),
#   
#   # filters define a restriction on the datasets
#   filters = 'hgnc_symbol', 
#   values = 'TH',
#   mart = humanMart)

# to retrieve all the genes
humanGenes = getBM(
  # values we want to retrieve
  attributes = c('entrezgene','hgnc_symbol','affy_hugene_1_0_st_v1'),
  mart = humanMart)

# write table
write.table(humanGenes,file = "../data/02.refs/humanGenes.txt",sep="\t",row.names = F,quote = F)
```